---
title: "NYPD_Shooting_Incident_Analysis"
date: "`r Sys.Date()`"
output: rmarkdown::github_document
---

* Questions of interest:
  + Which boroughs were most affected by the shootings? 
  + What were the number of shooting incidents by race and which race were affected the most and which the least?
  + What is the correlation between shootings by race and those that lead to deaths/murders?
  
* Description of NYPD_Shooting_Data
  + Each entity represents a single shooting incident for a single victim.
  + There are many fields included in the raw data many of which did not play a role in this analysis. Some of the fields that contribute valuable information for this analysis include: BORO (location of incident), STATISTICAL_MURDER_FLAG (whether the incident lead to murder), PERP_SEX, PERP_RACE, VIC_SEX, and VIC_RACE.

## Visualization #1 
* The following visualization compares the number of shooting incidents in boroughs in NYC. 
* It also compares the number of cases which lead to murders by using the STATISTICAL_MURDER_FLAG attribute.
* 'True" represents the number of shootings leading to death.
```{r, echo=TRUE}
##NYPD Shooting Data Import
library(tidyverse)
url_in <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

nypd_shooting_data <- read_csv(url_in)

##Tidy up - delte columns not needed, change data type for OCCUR_DATE
library(lubridate)
nypd_shooting_data <- nypd_shooting_data %>%
  select(-c(INCIDENT_KEY, X_COORD_CD, Y_COORD_CD, Latitude, Longitude, Lon_Lat,LOC_OF_OCCUR_DESC, PRECINCT, JURISDICTION_CODE, LOC_CLASSFCTN_DESC))

nypd_shooting_data <- nypd_shooting_data %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))

  
##See summary and filter through data as needed
summary(nypd_shooting_data)

##Visualization #1 Which borough has the most shootings and which has the least, and which of those
##lead to murders? Drop any rows with missing information on murder.
nypd_shooting_data %>%
drop_na(STATISTICAL_MURDER_FLAG) %>%
ggplot(aes(BORO, fill = STATISTICAL_MURDER_FLAG))+
geom_bar(alpha = 0.5)+
ggtitle("Shootings by Borough")+
theme(plot.title = element_text(hjust=0.5))
```

## Visualization #2 
* Next, we want to visualize the race of victims to see the race most frequently victimized.
```{r, echo=TRUE}

nypd_shooting_data %>%
drop_na(VIC_RACE) %>%
ggplot(aes(VIC_RACE))+
geom_bar(alpha = 0.5)+
labs(x = "Race of Victim", y = "Count")+
ggtitle("Race of Victims")+
theme(axis.text.x = element_text(size=6, angle = 15), 
      plot.title = element_text(hjust=0.5))

```

* The following pie charts reveal additional intriguing details about the shooting data:
  + "Victims from Shootings by Race" compares the total number of victims of shootings by race
  + "Confirmed Deaths from Shootings by Race" compares the total number of confirmed deaths from those shootings by race

* As expected, the two charts look almost identical as the chance of death should be similar when shot no matter the race of victim.

```{r, echo=TRUE}

slices <- c(sum(nypd_shooting_data$VIC_RACE == "AMERICAN INDIAN/ALASKAN NATIVE") 
                + sum(nypd_shooting_data$VIC_RACE == "UNKNOWN"),
            sum(nypd_shooting_data$VIC_RACE == "ASIAN / PACIFIC ISLANDER"),
            sum(nypd_shooting_data$VIC_RACE == "BLACK"),
            sum(nypd_shooting_data$VIC_RACE == "BLACK HISPANIC"),
            sum(nypd_shooting_data$VIC_RACE == "WHITE"),
            sum(nypd_shooting_data$VIC_RACE == "WHITE HISPANIC"))
lbls <- c("OTHER RACE", "ASIAN/P. ISLANDER", "BLACK", "BLK HISP.", "WHITE", "WHITE HISPANIC")
pie(slices, labels = lbls, main="Victims from Shootings by Race", cex=0.5)

slices <- c(sum(nypd_shooting_data$VIC_RACE == "AMERICAN INDIAN/ALASKAN NATIVE" & 
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE')
            +sum(nypd_shooting_data$VIC_RACE == "UNKNOWN" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'),
            sum(nypd_shooting_data$VIC_RACE == "ASIAN / PACIFIC ISLANDER" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'),
            sum(nypd_shooting_data$VIC_RACE == "BLACK" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'),
            sum(nypd_shooting_data$VIC_RACE == "BLACK HISPANIC" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'),
            sum(nypd_shooting_data$VIC_RACE == "WHITE" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'),
            sum(nypd_shooting_data$VIC_RACE == "WHITE HISPANIC" &
                  nypd_shooting_data$STATISTICAL_MURDER_FLAG == 'TRUE'))
lbls <- c("OTHER RACE", "ASIAN/P. ISLANDER", "BLACK", "BLK HISP.", "WHITE", "WHITE HISPANIC")
pie(slices, labels = lbls, main="Confirmed Deaths from Shootings by Race", cex = 0.5)
```

* Bias Identification
  + The NYPD shooting data exposes a disturbing reality of disproportionate impacts on the black community, where both the number of victims and confirmed deaths are alarmingly high. This outcome aligns with the pervasive systemic issues deeply rooted in our society. White Hispanics, black Hispanics, and other racial groups also experience significant disparities.

  + It's important to confront biases in data collection, including potential underreporting and limitations in racial categorization. As the researcher, I acknowledge the need for constant self-awareness and have strived to approach the analysis with a commitment to social justice.
  
  
  